Estimating SARS-CoV-2 transmission in England from randomly sampled prevalence surveys

### 2023-11-14 SEIR.tex

\todo[inline]{Add an intro with an overview of what this chapter contains. Include why I discuss mechanistic models, how I will use them, and what they add to the previous analyses.}

Mechanistic models explicitly represent the underlying processes driving the spread of infectious disease.
Infectious diseases have inherent feedback loops: current infectious individuals generate future infectious individuals.
Explicitly representing this system can generate insights not possible with the statistical methods used in the previous chapter.

The first sections of this chapter give the background and methodology of the model I use.
The literature on mechanistic models is vast, going back to \textcite{kermackContribution}.
Here, I explain only the background necessary to understand the model used in this chapter.
For further details I recommend the tutorial paper \textcite{kretzschmarMathematical} or the textbook \textcite{keelingModeling}.

@@

This section gradually builds complexity into a mechanistic model.
The mechanistic models considered here are \emph{compartmental} models.
A compartmental model divides all members of a population into compartments, each representing a disease state.
Compartmental models are the most common type of mechanistic infectious disease model.
They can be \emph{deterministic} or \emph{stochastic}.

A compartmental model consists of a state vector $\vec{x}(t)$ and a set of parameters $\vec{\theta}$.
The state vector is the proportion of the population (or number of individuals) in each compartment at time $t$.

Deterministic models have the property that, for any $t$, the state vector $\vec{x}(t)$ is a deterministic function of the parameters $\vec{\theta}$ and the initial state $\vec{x}(0)$.
The transmission model corresponds to the infection process (see \cref{E-inc-prev:sec:infection-process}).
There remains randomness in the observation model (described in \cref{SEIR:sec:observation}).

The other type of model is \emph{stochastic}.
Stochastic models specify a probability distribution $p(\vec{x} \mid \vec{\theta})$ for the state vector $\vec{x}(t)$ at time $t$.
Unlike the models previously encountered in this thesis, there will be dependence between $\vec{x}(t)$ and $\vec{x}(t')$.
For example, the number of individuals infected on day $t+1$, conditional on the model state on day $t$, might be a Poisson distribution.
The mean of the distribution is a function of the model state on day $t$.

I will use a deterministic model mathematically described by a system of Ordinary Differential Equations (ODE).
The derivative is taken with respect to time, $t$.
Therefore, in an ODE model, time is necessarily continuous although discretised when solving (see \cref{SEIR:sec:inference-implementation}).
For concreteness, I use units of days throughout, although the model can use any unit of time.
The number or proportion of individuals in each compartment is also continuous, which is a reasonable approximation when the population is large.
Equivalent formulations exist for the proportion or number of individuals.

When the number of infected individuals is large, then deterministic models are a good approximation to stochastic models.
Deterministic models are also known as \emph{mean-field} models because they can be derived from stochastic models by taking the mean of the distribution of the number of infected individuals.
Further discussion of stochastic models is beyond the scope of this thesis; \textcite[chapter 6]{keelingModeling} provides an introduction to these models.

\subsection{The SIR model} \label{SEIR:sec:SIR}
\begin{figure}[h]
\centering
\begin{tikzpicture}[
    node distance = 2.5cm,
    on grid,
    auto,
    ->,>=stealth',
    every state/.style={draw,rectangle,very thick},
    ]

@@

The simplest compartmental model is the \emph{susceptible-infected-recovered} (SIR) model (REF?).
The name refers to the three compartments in the model, displayed in \cref{SEIR:fig:SIR}.
Individuals in the susceptible compartment can be infected.
Individuals in the infected compartment can infect others, and eventually recover.
Individuals in the recovered compartment are immune to the disease, and cannot be infected again.

The SIR model is described by the following system of ODEs~\autocite[19]{keelingModeling}:
\begin{align}
\frac{ds(t)}{dt} &= -\beta s(t)i \\
\frac{di(t)}{dt} &= \beta s(t)i(t) - \gamma i(t) \\
\frac{dr(t)}{dt} &= \gamma i(t)
\end{align}
$s(t)$, $i(t)$ and $r(t)$ are the proportions of the population in the susceptible, infected and recovered compartments respectively.
These are the state of the system.
The state is a function of time, $t$, although this is not shown explicitly.
Two parameters are present in this model: $\beta$ and $\gamma$.
$\beta$ is the effective? transmission rate, or the number of individuals an infected individual infects per day.
$\gamma$ is the recovery rate, or the proportion of infected individuals that recover per day; hence, $1/\gamma$ is the mean infectious period.
These parameters are assumed to be constant over time (this assumption is relaxed in \cref{SEIR:sec:time-varying-foi}).

\todo[inline]{double check correct terms for the below, especially rates vs instantaneous proportion and per unit time vs per day}
The term $\beta s(t)i(t)$ is the rate at which susceptible individuals (as a proportion of the total population size) are infected~\autocite[18]{keelingModeling}.
Susceptible individuals move from the $s(t)$ to $i(t)$ when they are infected.
This term can be arrived at by considering a single infected individual.
Assume this individual makes contact with any other individual in the population at a constant rate, $\kappa$, in units of contacts per day.
Further, assume the population is \emph{well-mixed}, meaning the individual is equally likely to come into contact with any other individual.
For a randomly selected individual, the probability that they are susceptible is $s(t)$.
Each of the infected individual's contacts is therefore with a susceptible individual with probability $s(t)$, by the well-mixed assumption.
Therefore, in expectation, the infected individual makes contact with susceptible individuals at a rate of $\kappa s(t)$.
Assume that there is a constant probability of transmission upon contact between an infected and susceptible individual, $p_{si}$.
The expected rate of infections caused by this infected individual is therefore $\kappa s p_{si}$.
The two constant terms here (number of contacts and probability of transmission) are absorbed into the transmission rate, $\beta = \kappa p_{si}$.
Therefore, the instantaneous number of infections per infected individual is $\beta s(t)$.
The overall proportion of the population instantaneously infected is $\beta s(t)i(t)$.

@@

Between-age-group contacts rates have previously been inferred from a combination of contact surveys, time use data, and mobility data~\autocites{vanleeuwenTime}{vanleeuwenAugmenting}.
I use the outputs from this analysis as $\matr{K}$ in equation\todo{ref where this contact matrix fits in}.
To summarise, baseline rates and contexts of contacts between age groups are taken from the contact survey POLYMOD~\autocite{mossongSocial}.
How these contacts relate to time use is taken from the UK Time Use Survey\todo{cite time use survey}.
How the population spent their time each week during the pandemic is taken from the Google mobility data\todo{cite mobility data}, which measured the locations of Android users over the period of interest.
\todo[inline]{is the above paragraph too hand wavy?}

@@

Some behavioural changes (\eg mask wearing) and their consequent impacts on transmission are not captured by the contact survey.
To allow for these changes, $\beta$ is allowed to vary over time.
In particular, $\beta$ could? follows a random walk in log space on the weekly timescale:
\begin{align}
    \beta(t) &= \beta_{w(t)} \\
    \log\beta_{w+1} &= \log\beta_w + \sigma_\epsilon \epsilon_w \\
    \epsilon_w &\sim \N(0, 1)
\end{align}
where $w(t)$ is the week number of time $t$, with week 1 being the first week modelled.
Two parameters are unspecified here: the initial random of the random walk, $\beta_1$, and the standard deviation of the random walk, $\sigma_\epsilon$.
See \cref{perf-test:sec:parameters-priors} for the implementation and parameterisation of these.

>>>>

\todo[inline]{Add an intro with an overview of what this chapter contains. Include why I discuss mechanistic models, how I will use them, and what they add to the previous analyses. Differentiate the transmission model from the observation model.}

Mechanistic models explicitly represent the infection process driving the spread of infectious disease.
Infectious diseases have inherent feedback loops: current infectious individuals generate future infectious individuals.
Explicitly representing this system can generate insights not possible with the statistical methods used in the previous chapter\todo{give explicit examples of advantages}.

The first sections of this chapter give the background and methodology of the model I use.
The literature on mechanistic models is vast, a major early contribution being \textcite{kermackContribution}.
Here, I explain only the background necessary to understand the model used in this chapter.
For further details I recommend the tutorial paper \textcite{kretzschmarMathematical} or the textbook \textcite{keelingModeling}.

@@

\todo[inline]{integrate this paragraph into the intro part}
This section gradually builds complexity into a mechanistic model.
The class of mechanistic models considered here is \emph{compartmental} models, the most common type of mechanistic infectious disease model.
A compartmental model divides all members of the population under study into compartments, each representing a disease state.
They can be \emph{deterministic} or \emph{stochastic}.

A compartmental model consists of a state vector $\vec{x}(t)$ and a set of parameters $\vec{\theta}$.
The state vector is the proportion of the population (or number of individuals) in each compartment at time $t$.
Deterministic models have the property that, for any $t$, the state vector $\vec{x}(t)$ is a deterministic function of the parameters $\vec{\theta}$ and the initial state $\vec{x}(0)$.
\todo{Maybe say something about still having a stochastic observation model, if not clear enough from intro}
%The transmission model corresponds to the infection process (see \cref{E-inc-prev:sec:infection-process}).
%There remains randomness in the observation model (described in \cref{SEIR:sec:observation}).
The other type of model is \emph{stochastic}.
Stochastic models specify a probability distribution $p(\vec{x} \mid \vec{\theta})$ for the state vector $\vec{x}(t)$ at time $t$.
Unlike a Poisson process (see\todo{ref intro section}), there is a dependence between $\vec{x}(t)$ and $\vec{x}(t')$.
Further discussion of stochastic models is beyond the scope of this thesis; see \textcite[chapter 6]{keelingModeling} for an introduction.

This chapter focuses on deterministic models mathematically described by a system of ODEs, specifically $d\vec{x}/dt$.
Therefore, in an ODE model, time is necessarily continuous, although discretised when solving (see \cref{SEIR:sec:inference-implementation}).
For concreteness, I use units of days throughout, although the model can use any unit of time.
The number or proportion of individuals in each compartment is also continuous, which is a reasonable approximation when the population is large.

When the number of infected individuals is large, then deterministic models are a good approximation to stochastic models.
Deterministic models are also known as \emph{mean-field} models because they can be derived from stochastic models by taking the mean of the distribution of the number of infected individuals.

\subsection{The SIR model} \label{SEIR:sec:SIR}

\begin{figure}[h]
\centering
\begin{tikzpicture}[
    node distance = 2.5cm,
    on grid,
    auto,
    ->,>=stealth',
    every state/.style={draw,rectangle,very thick},
    ]

@@

The simplest compartmental model is the \emph{susceptible-infected-recovered} (SIR) model\todo{ref for SIR being simplest}.
The name refers to the three compartments in the model, displayed in \cref{SEIR:fig:SIR}.
Individuals in the susceptible compartment can be infected.
Individuals in the infected compartment can infect others, and eventually recover.
Individuals in the recovered compartment are immune to the disease, and cannot be infected again.

The SIR model is described by the following system of ODEs~\autocite[19]{keelingModeling}:
\begin{align}
\frac{ds(t)}{dt} &= -\beta s(t)i \\
\frac{di(t)}{dt} &= \beta s(t)i(t) - \gamma i(t) \\
\frac{dr(t)}{dt} &= \gamma i(t)
\end{align}
$s(t)$, $i(t)$ and $r(t)$ are the proportions of the population in the susceptible, infected and recovered compartments respectively.
These are the state of the system.
The state is a function of time, $t$, although this is not shown explicitly.
Two parameters are present in this model: $\beta$ and $\gamma$.
$\beta$ is the transmission rate\todo{check what beta is called: effective transmission rate or transmission rate or something else?}, or the number of individuals an infected individual infects per day.
$\gamma$ is the recovery rate, or the proportion of infected individuals that recover per day; hence, $1/\gamma$ is the mean infectious period.
These parameters are assumed to be constant over time (this assumption is relaxed in \cref{SEIR:sec:time-varying-foi}).

\todo[inline]{double check correct terms for the below, especially rates vs instantaneous proportion and per unit time vs per day}
The term $\beta s(t)i(t)$ is the rate at which susceptible individuals (as a proportion of the total population size) are infected~\autocite[18]{keelingModeling}.
Susceptible individuals move from the $s(t)$ to $i(t)$ when they are infected.
This term can be arrived at by considering a single infected individual.
Assume this individual makes $\kappa$ contacts with individuals in the population per day.
Further, assume the population is \emph{well-mixed}, meaning the each contact is with an individual chosen uniformly at random from the population.
For such a randomly selected individual, the probability that they are susceptible is $s(t)$.
Each of the infected individual's contacts is therefore with a susceptible individual with probability $s(t)$, by the well-mixed assumption.
Therefore, in expectation, the infected individual makes contact with susceptible individuals at a rate of $\kappa s(t)$.
Assume that there is a constant probability of transmission upon contact between an infected and susceptible individual, $p_{si}$.
The expected rate of infections caused by this infected individual is therefore $\kappa s p_{si}$.
The two constant terms here (number of contacts and probability of transmission) are absorbed into the transmission rate, $\beta = \kappa p_{si}$.
Therefore, the instantaneous number of infections per infected individual is $\beta s(t)$.
The overall proportion of the population instantaneously infected is $\beta s(t)i(t)$.

@@

Between-age-group contacts rates have previously been inferred from a combination of contact surveys, time use data, and mobility data~\autocites{vanleeuwenTime}{vanleeuwenAugmenting}.
I use the outputs from this analysis as $\matr{K}$ in equation\todo{ref where this contact matrix fits in}.
To summarise, baseline rates and contexts of contacts between age groups are taken from the contact survey POLYMOD~\autocite{mossongSocial}.
How these contacts relate to time use is taken from the UK Time Use Survey\todo{cite time use survey}.
How the population spent their time each week during the pandemic is taken from the Google mobility data\todo{cite mobility data}, which measured the locations of Android users over the period of interest.
\todo[inline]{add a bit more detail to the above}

@@

Some behavioural changes (\eg mask wearing) impact transmission.
However, these are not captured by the contact survey.
To allow for these changes, $\beta$ is allowed to vary over time.
In particular, $\beta$ follows a random walk in log space on the weekly timescale:
\begin{align}
    \beta(t) &= \beta_{w(t)} \\
    \log\beta_{w+1} &= \log\beta_w + \sigma_\epsilon \epsilon_w \\
    \epsilon_w &\sim \N(0, 1)
\end{align}
where $w(t)$ is the week number of time $t$, with week 1 being the first week modelled.
Two parameters are unspecified here: the initial random of the random walk, $\beta_1$, and the standard deviation of the random walk, $\sigma_\epsilon$.
See \cref{perf-test:sec:parameters-priors} for the implementation and parameterisation of these.
